Concurrent Programming in Harmony

### 2021-12-24 book/paper.tex
Fixes to book

Harmony programs are compiled to Harmony \emph{bytecode}
\index{bytecode}%
(a list of machine instructions for a virtual machine),
which in turn is executed by the Harmony virtual machine (HVM).
\index{virtual machine}%
\index{Harmony Virtual Machine}%
\index{HVM}%
The Harmony compiler places the bytecode for file <{x}>.\texttt{htm} in
file <{x}>.\texttt{hvm}.
The model checker, called \emph{Charm}, places its output in a file
called <{x}>.\texttt{hco} and a stylized version of the same output
in an HTML file called <{x}>.\texttt{htm}.

@@

Harmony programs, and indeed the HVM,  manipulate Harmony values.
Harmony values are recursively defined:
they include booleans (<{False}> and <{True}>),
integers (but not floating point numbers),
strings (enclosed by single or double quotes),
sets of Harmony values, and dictionaries
\index{dictionary}%
that map Harmony values to other Harmony values.
Strings that start with a letter or an underscore and only contain
letters, digits, and underscores can be written without quotes by
preceding it with a dot.  So \texttt{.example} is the same string
as \texttt{"example"}.

Harmony makes extensive use of dictionaries.
A dictionary maps keys to values.
Unlike Python, any Harmony value can be a key, including another
dictionary.
Dictionaries are written as
$\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If <{d}> is a dictionary, and <{k}> is a key, then the
following expression retrieves the Harmony value that <{k}> maps to in <{d}>:
\begin{code}
<{d k}>
\end{code}
The meaning of <{d a b}> $...$ is (<{((d a) b)}> $...$).
This notation is unfamiliar to Python programmers,
but in Harmony square brackets can be used
in the same way as parentheses, so you can express the same thing in the form
that is familiar to Python programmers:
\begin{code}
<{d[k]}>
\end{code}
However, if <{d = { .count: 3 }}>, then you can write
<{d.count}>
(which has value 3) instead of having to write
<{d[.count]}> or <{d["count"]}> (although any of those
will work).
Thus a dictionary can be made to look much like a Python object.

@@

\autoref{ap:values} provides details on all the values that
Harmony currently supports.

@@

Instead of bits and bytes, a HVM manipulates Harmony values.
A HVM has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of HVM instructions,
generated from a Harmony program.  The types of instructions will be described later.
\item Shared memory: A HVM has just one memory location containing
a Harmony value.
\item Threads:  Any thread
can spawn an unbounded number of other threads and threads may terminate.
Each thread has a program counter that indexes into the code,
\index{program counter}%
a stack of Harmony values,
and two private \emph{registers} that each contain a Harmony value.
\index{register}%
\end{itemize}

\glsadd{context}%
\index{thread-local}%
One of the registers of a thread contains the local variables of the
method that the thread is currently executing.  It is saved and restored by
method invocations.  The other register is called \textbf{this} and is
used for so-called \emph{thread-local state}, which contains variables that
can be used by all methods.
The state of a thread is called a \emph{context} (aka \emph{continuation}):
\index{context}%
\index{continuation}%
it contains the values of its program counter, stack, and registers.
The state of a HVM
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}%
\index{multiset}%
of contexts.  It is a multiset of contexts because two threads can
have the same context.

@@

It may seem strange that there is only one memory location.
However, this is not a limitation because Harmony values are unbounded trees.
The shared memory is a dictionary that maps strings (names of shared
variables) to other Harmony values.
We call this a \emph{directory}.
\index{directory}%
Thus, a directory represents the state of a collection of variables named
by the strings.
%
Because directories are Harmony values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of Harmony values, like a path name in the file system
hierarchy.  We call such a sequence an \emph{address}.
\index{address}%
For example, in \autoref{fig:inc} the memory is a dictionary with two
entries: \texttt{.count} and \texttt{.done}.  And the value of entry
\texttt{.done} is a dictionary with keys 0 and 1.
So, for example, the address of \texttt{done[0]} is the sequence
[ \texttt{.done}, 0 ].
An address is itself a Harmony value.

>>>>

Harmony programs are compiled to Harmony \emph{bytecode}
\index{bytecode}%
(a list of machine instructions for a virtual machine),
which in turn is executed by the Harmony virtual machine (HVM).
\index{virtual machine}%
\index{Harmony Virtual Machine}%
\index{HVM}%
The Harmony compiler places the bytecode for file <{x}>.\texttt{htm} in
file <{x}>.\texttt{hvm}.
The model checker, called \emph{Charm}, executes the code in
<{x}>.\texttt{htm} and places its output in a file
called <{x}>.\texttt{hco} and a stylized version of the same output
in an HTML file called <{x}>.\texttt{htm}.

@@

Harmony programs, and indeed the HVM,  manipulate Harmony values.
Harmony values are recursively defined:
they include booleans (<{False}> and <{True}>),
integers (but not floating point numbers),
strings (enclosed by single or double quotes),
sets of Harmony values, and dictionaries
\index{dictionary}%
that map Harmony values to other Harmony values.
Strings that start with a letter or an underscore and only contain
letters, digits, and underscores can be written without quotes by
preceding it with a dot.  So <{.example}> is the same string
as <{"example"}>.

Harmony makes extensive use of dictionaries.
A dictionary maps keys to values.
Unlike Python, which requires that keys must be hashable,
any Harmony value can be a key, including another dictionary.
Dictionaries are written as
$\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If <{d}> is a dictionary, and <{k}> is a key, then the
following expression retrieves the Harmony value that <{k}> maps to in <{d}>:
\begin{code}
<{d k}>
\end{code}
The meaning of <{d a b}> $...$ is (<{((d a) b)}> $...$).
This notation is unfamiliar to Python programmers,
but in Harmony square brackets can be used
in the same way as parentheses, so you can express the same thing in the form
that is familiar to Python programmers:
\begin{code}
<{d[k]}>
\end{code}
However, if <{d = { .count: 3 }}>, then you can write
<{d.count}>
(which has value 3) instead of having to write
<{d[.count]}> or <{d["count"]}> (although any of those
will work).
Thus a dictionary can be made to look much like a Python object.

@@

\autoref{ap:values} provides details on all the types of values that
Harmony currently supports.

@@

\index{register}%
\glsadd{context}%
\index{thread-local}%
Instead of bits and bytes, a HVM manipulates Harmony values.
A HVM has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of HVM instructions,
generated from a Harmony program.  The types of instructions will be described later.
\item Shared memory: A HVM has just one memory location containing
a Harmony value.
\item Threads:  Any thread
can spawn an unbounded number of other threads and threads may terminate.
Each thread has a program counter that indexes into the code,
\index{program counter}%
a stack of Harmony values,
and a private \emph{register} that contains a
Harmony value.\footnote{Currently, another thread register
contains thread-local data.  We do not use it (yet) in this book.}
\end{itemize}

The register of a thread contains the local variables of the
method that the thread is currently executing.  It is saved and restored by
method invocations.
The state of a thread is called a \emph{context} (aka \emph{continuation}):
\index{context}%
\index{continuation}%
it contains the values of its program counter, stack, and registers.
The HVM state
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}%
\index{multiset}%
of contexts.  It is a multiset of contexts because two threads can
have the same context at the same time.

@@

It may seem strange that there is only one memory location.
However, this is not a limitation because Harmony values are unbounded trees.
The shared memory is a dictionary that maps strings (names of shared
variables) to other Harmony values.
We call this a \emph{directory}.
\index{directory}%
Thus, a directory represents the state of a collection of variables named
by the strings.
%
Because directories are Harmony values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of Harmony values, like a path name in the file system
hierarchy.  We call such a sequence an \emph{address}.
\index{address}%
For example, in \autoref{fig:inc} the memory is a dictionary with two
entries: <{.count}> and <{done}>.  And the value of entry
<{done}> is a dictionary with keys 0 and 1.
So, for example, the address of <{done[0]}> is the sequence <{[done, 0]}>.
An address is itself a Harmony value.
